Rhizaria in the oligotrophic exhibit clear temporal and vertical variability.

### 2024-01-25 manuscript/02_methods.qmd
finalize figs for draft

Data were collected in collaboration with the Bermuda Atlantic Time-Series Study [@michaels1996; @lomas2013] on board the R/V Atlantic Explorer. Cruises were conducted at approximately monthly intervals. Rhizaria individuals were sampled using the Underwater Vision Profiler 5 [UVP5; @picheral2010], a tool which is well established to accurately quantify large Rhizaria [@biard2016; @stukel2018; @stukel2019; @biard2020; @barth2022; @drago2022; @llopismonferrer2022; @panaïotis2023]. The UVP5 was mounted to the sampling rosette and collected data autonomously on routine casts, from which only the downcast data are utilized. The UVP5 was deployed from June-September 2019 then from October 2020 - January 2022. Casts were filtered to only include data collected in the BATS region, far offshore of Bermuda into the Sargasso Sea (Supplemental Figure 1; Supplemental Table 1). In general, casts extended to either 200m, 500m, or 1200m deep, with a few extended into the bathypelagic (4500m). However, Rhizaria were only typically found in large abundances throughout the epipelagic and mesopelagic zones. As such, we limit this study to results from the upper 1000m of the water column.

A variety of biotic and abiotic data were collected during each BATS cruise. Briefly, we will explain the data utilized in this study. Salinity, temperature, Dissolved Oxygen (DO), and in-situ chlorophyll fluorescence were measured at high-frequencies on each cast, attached to the same CTD rosette as the UVP. On certain casts, niskin bottles were used to collect bacterial abundance estimates (via epifluorescence microscopy) as well as measure inorganic nutrients ($NO_3$, and $Si$ as silicate/sicilic acid) at discrete depths. On each cruise, flux estimates of total mass, carbon, and nitrogen were also collected using sediment traps (in the present study we utilize flux to the mesopelagic as the flux at 200m). Also primary productivity was estimated through measuring $C^{14}$ uptake rates from in-situ incubations. Full descriptions of the BATS sampling program, methods, and data are available freely online (https://bats.bios.asu.edu/bats-data/) and reviewed in [@lomas2013].

@@

Individual vignettes of Rhizaria images were identified using the classification platform Ecotaxa [cite]. Data were pre-sorted according to a . Taxonomic classification were done based on morphology exclusively. While there are sparse taxonomic guides for in-situ images of rhizaria, identification largely relied on descriptions in [@nakamura2015; @suzuki2015; and @biard2020]. Using the aforementioned sources and publicly available ecotaxa projects, I constructed a guide which can be found here: (https://thealexbarth.github.io/media//Project_Items/Oligotrophic_Community/ecotaxa_UVP-guide-stone-lab.pdf). Broadly, Rhizaria were classified as Foraminifera, Radiolarians (Acantharea or Collodaria), or a variety of Phaeodarian families (Figure 1). When identification could not be confidently made between a few candidate taxa, a less specific label was used. As a result, we have data from "unidentified Rhizaria", which typically were vignettes not distinguishable between Aulacanthidae or Acantharea or "unidentified Phaeodaria", which are clearly phaeodaria but not distinguishable into a family.

The UVP5 samples at \~15Hz rate as it descends the water column and records the exact position of each particle larger than 600$\mu m$. However, identified rhizaria ranged from a 934$\mu m$ Aulacanthidae cell to a Collodarian colony over 10mm in diameter. To confirm that the UVP5 was sampling adequately across all size ranges, an NBSS slope was constructed to identify a drop-off which would indicate poor-sampling at the small size range [@lombard2019; @barth-inrev] . However, it was evident from this analysis that all size ranges were adequately sampled across the size range (Supplemental Figure 2) so no data were excluded. The UVP5 reports the exact depth at which a particle is recorded, however to estimate abundance, observations must be binned over fixed depth intervals. In our deployments had variable descent depths and speeds with more casts descending to 500m than 1000m and descents quicker through the epipelagic than the mesopelagic [see @barth2022 for an extended discussion of UVP5 data processing). For the present study, Rhizarian abundances were estimated in 25m vertical bins, which offer a moderate sampling volume (average 0.948$m^3$ in the epipelagic and 0.589$m^3$ in the mesopelagic) while still maintaining ecologically relevant widths. However, concentrations in a 25m bin would need to be greater than 2.428 indv. $m^{-3}$ and 3.912 indv. $m^{-3}$, in the epipealgic and mesopelagic respectively, to fall below a 10% non-detection risk [@benfield1996; @barth-inrev]. Because we typically observed many rhizaria taxa below these concentrations, we present the 25m binned data to visualize broad-scale average distributions. For quantifying and modelling Rhizaria abundances, we present integrated abundance estimates, with each cast. Due to the variable descent depths of the UVP, data are categorized as epipelagic (0-200m), upper mesopelagic (200-500m), and lower mesopelagic (500-1000m). The average sampling volume integrated through these regions were 7.59$m^2$, 7.06$m^2$, and 11.77$m^2$, with non-detection thresholds at 0.30 indv. $m^{-2}$, 0.33 indv. $m^{-2}$, and 0.20 indv. $m^{-2}$ respectively.

@@

Generalized Additive Models (GAMs) were used to assess the relationship between integrated rhizarian abudance and different environmental factors.

>>>>

Data were collected in collaboration with the Bermuda Atlantic Time-Series Study [@michaels1996; @lomas2013] on board the R/V Atlantic Explorer. Cruises were conducted at approximately monthly intervals. Rhizaria individuals were sampled using the Underwater Vision Profiler 5 [UVP5; @picheral2010], a tool which is well established to accurately quantify large Rhizaria [@biard2016; @stukel2018; @stukel2019; @biard2020; @barth2022; @drago2022; @llopismonferrer2022; @panaïotis2023]. The UVP5 was mounted to the sampling rosette and collected data autonomously on routine casts, from which only the downcast data are utilized. The UVP5 was deployed from June-September 2019 then from October 2020 - January 2022. Casts were filtered to only include data collected in the BATS region, far offshore of Bermuda into the Sargasso Sea (Supplemental Figure 1). In general, casts extended to either 200m, 500m, or 1200m deep, with a few extended into the bathypelagic (4500m). However, Rhizaria were only typically found in large abundances throughout the epipelagic and mesopelagic zones. As such, we limit this study to results from the upper 1000m of the water column.

A variety of biotic and abiotic data were collected during each BATS cruise. Briefly, we will explain the data utilized in this study. The UVP5 provided particle count data at a high-frequency from each cast. Particle concentration was calculated from this data for all particles 184$\mu m$ - 450$\mu m$. The lower size range was set by what could be reliably sampled by the UVP5's pixel resolution (\>2px) and the upper size range is representative of a potential prey field for mesozooplankton [@whitmore2021]. Salinity, temperature, Dissolved Oxygen (DO), and in-situ chlorophyll fluorescence were measured at high-frequencies on each cast, attached to the same CTD rosette as the UVP. On certain casts, niskin bottles were used to collect bacterial abundance estimates (via epifluorescence microscopy) as well as measure inorganic nutrients ($NO_3$, and $Si$ as silicate/sicilic acid) at discrete depths. On each cruise, flux estimates of total mass, carbon, and nitrogen were also collected using sediment traps (in the present study we utilize flux to the mesopelagic as the flux at 200m). Also primary productivity was estimated through measuring $C^{14}$ uptake rates from in-situ incubations. Full descriptions of the BATS sampling program, methods, and data are available freely online [(https://bats.bios.asu.edu/bats-data/)](https://bats.bios.asu.edu/bats-data/) and reviewed in [@lomas2013].

@@

Individual vignettes of Rhizaria images were identified using the classification platform Ecotaxa \[cite\]. Data were pre-sorted according to a . Taxonomic classification were done based on morphology exclusively. While there are sparse taxonomic guides for in-situ images of rhizaria, identification largely relied on descriptions in [@nakamura2015; @suzuki2015; and @biard2020]. Using the aforementioned sources and publicly available ecotaxa projects, I constructed a guide which can be found here: <https://thealexbarth.github.io/media//Project_Items/Oligotrophic_Community/ecotaxa_UVP-guide-stone-lab.pdf>. Broadly, Rhizaria were classified as Foraminifera, Radiolarians (Acantharea or Collodaria), or a variety of Phaeodarian families (Figure 1). When identification could not be confidently made between a few candidate taxa, a less specific label was used. As a result, we have data from "unidentified Rhizaria", which typically were vignettes not distinguishable between Aulacanthidae or Acantharea or "unidentified Phaeodaria", which are clearly phaeodaria but not distinguishable into a family.

![Example images of different Rhizaria taxa. 4mm scale bar shown in lower right. All vignettes are same scale](images/01_taxa.pdf)

The UVP5 samples at \~15Hz rate as it descends the water column and records the exact position of each particle larger than 600$\mu m$. However, identified rhizaria ranged from a 934$\mu m$ Aulacanthidae cell to a Collodarian colony over 10mm in diameter. To confirm that the UVP5 was sampling adequately across all size ranges, an NBSS slope was constructed to identify a drop-off which would indicate poor-sampling at the small size range [@lombard2019; @barth-inrev] . However, it was evident from this analysis that all size ranges were adequately sampled across the size range (Supplemental Figure 2) so no data were excluded. The UVP5 reports the exact depth at which a particle is recorded, however to estimate abundance, observations must be binned over fixed depth intervals. In our deployments had variable descent depths and speeds with more casts descending to 500m than 1000m and descents quicker through the epipelagic than the mesopelagic [see @barth2022 for an extended discussion of UVP5 data processing). For the present study, Rhizarian abundances were estimated in 25m vertical bins, which offer a moderate sampling volume (average 0.948$m^3$ in the epipelagic and 0.589$m^3$ in the mesopelagic) while still maintaining ecologically relevant widths. However, concentrations in a 25m bin would need to be greater than 2.428 indv. $m^{-3}$ and 3.912 indv. $m^{-3}$, in the epipelagic and mesopelagic respectively, to fall below a 10% non-detection risk [@benfield1996; @barth-inrev]. Because we typically observed many rhizaria taxa below these concentrations, we present the 25m binned data to visualize broad-scale average distributions. For quantifying and modelling Rhizaria abundances, we present integrated abundance estimates, with each cast. Due to the variable descent depths of the UVP, data are categorized as epipelagic (0-200m), upper mesopelagic (200-500m), and lower mesopelagic (500-1000m). The average sampling volume integrated through these regions were 7.59$m^2$, 7.06$m^2$, and 11.77$m^2$, with non-detection thresholds at 0.30 indv. $m^{-2}$, 0.33 indv. $m^{-2}$, and 0.20 indv. $m^{-2}$ respectively. All UVP data processing was done using the `EcotaxaTools` package in R (Barth 2023).

@@

Generalized Additive Models (GAMs) were used to assess the relationship between integrated Rhizarian abundance and different environmental factors. GAMs offer the ability to model non-linear and non-monotonic relationships, which can be particularly useful in assessing ecological relationships [@wood2017] and have been successfully applied to Rhizarian ecology [@biard2020]. The `mgcv` package [@wood2001] was used to construct models relating environmental parameters to each taxonomic group's integrated abundance estimates from each cast. To select the most parsimonious model for each analysis, a backwards step-wise approach was taken. First, a full model was fit using any term which may be ecologically relevant. Terms were fit using maximum likelihood with a double penalty approach on unnecessary smooths [@marra2011]. The smoothness parameter was restricted (k = 6) to prevent overfitting the models. At each iteration of the backwards step-wise procedure, the model term with the lowest F score (least statistically significant) was removed. This was repeated until all model terms were statistically significant or the $R^2_{adj}$ was substantially reduced. Models were fit for each region; epipelagic, upper mesopelagic, and lower mesopelagic. In cases where observations were too sparse for a given taxonomic grouping, models were not run.

### 2024-01-25 manuscript/main.qmd
finalize figs for draft

---
title: Rhizaria in the oligotrophic exhibit clear temporal and vertical variability.
bibliography: references.bib
csl: ./deep-sea-research-part-i.csl
execute:
  echo: false
  warning: false
  message: false
author:
  - name: Alex Barth*
    affiliations:
    - name: University of South Carolina
      department: Biological Sciences
      address: 700 Sumter St 401, Columbia, SC 29208
      email: AB93@email.sc.edu
      phone: +1 (440) 668-8376
  - name: Leo Blanco-Berical
    affiliations:
    - name: Arizone State University
  - name: Rod Johnson
    affiliations: 
    - name: Bermuda Institute of Ocean Sciences
      address: Bermuda GE 01, UK
  - name: Joshua Stone
    affiliations:
    - name: University of South Carolina
      department: Biological Sciences
      address: 700 Sumter St 401, Columbia, SC 29208
format:
  pdf:
    # template-partials:
    #   - "before-body.tex"
    #   - "_titlepage.tex"
    # include-in-header:
    #   - "in-header.tex"
    number-sections: false
    linestretch: 2
    tbl-cap-location: top
    # filters: [wordcount.lua]
    keep-tex: true
---

@@

>>>>

---
title: Rhizaria in the oligotrophic exhibit clear temporal and vertical variability.
bibliography: references.bib
csl: ./deep-sea-research-part-i.csl
execute:
  echo: false
  warning: false
  message: false
author:
  - name: Alex Barth*
    affiliations:
    - name: University of South Carolina
      department: Biological Sciences
      address: 700 Sumter St 401, Columbia, SC 29208
      email: AB93@email.sc.edu
      phone: +1 (440) 668-8376
  - name: Leo Blanco-Berical
    affiliations:
    - name: Arizone State University
  - name: Rod Johnson
    affiliations: 
    - name: Bermuda Institute of Ocean Sciences
      address: Bermuda GE 01, UK
  - name: Joshua Stone
    affiliations:
    - name: University of South Carolina
      department: Biological Sciences
      address: 700 Sumter St 401, Columbia, SC 29208
format:
  pdf:
    template-partials:
      - "before-body.tex"
      - "_titlepage.tex"
    include-in-header:
      - "in-header.tex"
    number-sections: false
    linestretch: 2
    tbl-cap-location: top
    filters: [wordcount.lua]
    keep-tex: true
---

@@

```{r}
#| child: ./02_methods.qmd
```

```{r}
#| child: ./03_results.qmd
```

```{r}
#| child: ./04_discussion.qmd
```

# References:
:::{#refs}
:::
